---
title: "RMHI/ARMP Problem Set 2"
author: 'Your ID goes here [Word Count: XX]'
output: word_document
---

Please put your answers here, following the instructions in the assignment description. Put your answers and word count tallies in the locations indicated; if none is indicated that means there is no word count for that question. Remember to knit as you go, and submit the knitted version of this on Canvas.

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
# We'll begin by loading up the libraries and data we need, as always.
knitr::opts_chunk$set(echo = TRUE)

# this deletes any variables that are in your environment
# useful so you don't get conflicts you're unaware of
rm(list=ls())

# loading the libraries
library(tidyverse)
library(here)
library(ggplot2)
# you might need other libraries; load here here if so


# loading datasets (don't change this)
ds <- read_csv(file=here("sneezes.csv"))
dl <- read_csv(file=here("liedetector.csv"))
df <- read_csv(file=here("foodwaste.csv"))
da <- read_csv(file=here("animals.csv"))
dh <- read_csv(file=here("health.csv"))
dp <- read_csv(file=here("positivity.csv"))

# makes variables into factors (don't change this)
dh$person <- as.factor(dh$person)
dp$person <- as.factor(dp$person)
dl$question <- as.factor(dl$question)
ds$distCat <- as.factor(ds$distCat)

# reorders so orders are intuitive, and converts to factors
# (don't change this)
da$size <- factor(da$size,levels=c("small","medium","large"))
da$year <- factor(da$year,levels=c("past","present"))
dh$height <- factor(dh$height,levels=c("short","tall"))
dh$income <- factor(dh$income,levels=c("poor","rich"))
dp$then <- factor(dp$then,levels=c("positive","negative"))
dp$now <- factor(dp$now,levels=c("positive","negative"))
```


## Q1


**Q1a**

```{r q1a, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
f_five_data=head(ds)
cor.test(ds$distance,ds$sneezes)
```

*ANSWER:* XXX [Word Count: N]


**Q1b**

```{r q1b, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
library(ggplot2)
library(dplyr)

summary_df <- ds %>%
  group_by(distCat) %>%
  summarise(
    mean_sneezes = mean(sneezes),
    se = sd(sneezes) / sqrt(n())
  )

colors <- c("blue", "green", "red")
names(colors) <- levels(ds$distCat)

ggplot(summary_df, aes(x = distCat, y = mean_sneezes, fill = distCat)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.6) +
  geom_errorbar(aes(ymin = mean_sneezes - se, ymax = mean_sneezes + se),
                width = 0.2) +
  geom_point(data = ds, 
             aes(x = distCat, y = sneezes, color = distCat), 
             position = position_nudge(x = 0.05), 
             alpha = 0.8, size = 2, show.legend = FALSE) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  theme_minimal() +
  labs(
    title = "Mean Number of Sneezes by Distance Category",
    x = "Distance Category",
    y = "Mean Number of Sneezes"
  ) +
  theme(legend.position = "none")

```


**Q1c**

```{r q1c, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
library(car)

leveneTest(sneezes ~ distCat, data = ds)

model <- aov(sneezes ~ distCat, data = ds)
shapiro.test(residuals(model))

```

*Note that the order of these assumptions (i.e., what assumption you put in 1 vs 2) does not matter! The words ASSUMPTION 1 and ASSUMPTION 2 do not contribute to word count.* 

*ASSUMPTION 1:* XXX [Word Count: N]

*ASSUMPTION 2:* XXX [Word Count: N]


**Q1d**

```{r q1d, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
anova_model <- aov(sneezes ~ distCat, data = ds)
summary(anova_model)
```

*ANSWER:* XXX [Word Count: N]


**Q1e**

```{r q1e, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
TukeyHSD(anova_model)
```

*ANSWER:* XXX [Word count: N]


**Q1f**

*ANSWER:* XXX [Word count: N]


## Q2 


**Q2a** 

```{r q2a, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q2b** 

```{r q2b, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q2c** 

```{r q2c, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q2d** 

```{r q2d, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


## Q3


**Q3a**

```{r q3a, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


## Q4

**Q4a**

```{r q4a, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q4b**

```{r q4b, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* (i) XXX (ii) XXX [Word count: N]


**Q4c**

```{r q4c, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q4d**

```{r q4d, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* (i) XXX (ii) XXX [Word count: N]


## Q5


**Q5a**

```{r q5a, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q5b**

```{r q5b, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


## Q6


**Q6a**

```{r q6a, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q6b**

```{r q6b, warning=FALSE, message=FALSE, fig.width=4.5, fig.height=3}
# insert your code here
```

*ANSWER:* XXX [Word count: N]


**Q6c** 

*ANSWER:* XXX [Word count: N]


## Q7


**Q7a**

*ANSWER:* (i) XXX (ii) XXX (iii) XXX 


**Q7b**

*ANSWER:* (i) XXX (ii) XXX (iii) XXX [Word Count: N]


## Q8


*ANSWER:* XXX

